---
title: "MPG of Auto vs Manual Transmissions"
author: "kmok23"
date: "October 22, 2015"
output: html_document
---

##Executive Summary
"Is an automatic or manual transmission better for MPG"  
"Quantify the MPG difference between automatic and manual transmissions"

##Exploratory Data Analyses
A quick preview of the Motor Trends MPG data provides the gas mileage for 
various cars as well as other characteristics of the various cars.
```{r echo=FALSE}
data(mtcars)

head(mtcars)
```

A boxplot of the MPG data versus the transmission type shows that the manual 
transmission is better for MPG. An initial look shows the average MPG of 
automatic transmissions at `r round(mean(mtcars$mpg[mtcars$am == 0]),2)` and the
average MPG of manual transmissions at 
`r round(mean(mtcars$mpg[mtcars$am == 1]),2)`.

```{r echo=FALSE}
boxplot(mpg ~ am, data = mtcars, 
        main = "MPG by Transmission", 
        xlab = "Transmission", 
        ylab = "MPG", 
        names = c("Automatic", "Manual"))
```

##Regression Models

###Simple Linear Regression
The initial model tackles the main question with a linear model and am as the 
only regressor: $mpg = \beta_0 + \beta_1 am$. For this regression, 
$\beta_0 = 17.147$ and $\beta_1 = 7.245$.

###Multivariate Regression
A review of the correlation of the MPG variable to the other variables show that 
the cyl, disp, hp, and wt variables have the highest correlation with MPG. These 
variables should be considered for inclusion in the model. It is also noted that 
cyl has very high correlation with disp and hp as well.

Did the student interpret the coefficients correctly?
Did the student do some exploratory data analyses?
Did the student fit multiple models and detail their strategy for model selection?
Did the student answer the questions of interest or detail why the question(s) is (are) not answerable?
Did the student do a residual plot and some diagnostics?
Did the student quantify the uncertainty in their conclusions and/or perform an inference correctly?
Was the report brief (about 2 pages long) for the main body of the report and no longer than 5 with supporting appendix of figures?
Did the report include an executive summary?

\pagebreak

##Appendix

Correlation matrix for MPG data:
```{r}
cor(mtcars)
```

Basic linear model with am as regressor:
```{r}
mdl <- lm(mpg ~ am, data = mtcars)
summary(mdl)
```